\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-28 paper.qmd
done with first image

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "none", 
     legend.title = element_text(size=5), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
```

@@

```{r}
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")

nldr1 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1) +
  ggtitle("a")
```

```{r}
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")

nldr2 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("b")

```

```{r}
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.99.rds")
nldr3 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("c") 
```

```{r}
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")

nldr4 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("d")  
```

```{r}
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")

nldr5 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("e") 
```

```{r}
phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
nldr6 <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("f")  
```

```{r}
trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
nldr7 <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("g") 
```

```{r}
pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
nldr8 <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("h")
```

```{r}
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Which is the most accurate representation of the structure in high dimensions?"
#| label: fig-diffnldrpbmc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
# 2D layouts from different NLDR techniques and different hyperparameter choices applied for the PBMC3k dataset: (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (d) UMAP (n_neighbors = 15, min_dist = 0.03), (e) UMAP (n_neighbors = 84, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (h) tSNE (perplexity = 92), (i) PHATE (knn = 2), (j) PHATE (knn = 5), (k) PHATE (knn = 10), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (m) TriMAP (n_inliers = 30, n_outliers = 4, n_random = 10), (n) TriMAP (n_inliers = 5, n_outliers = 2, n_random = 2), (p) PaCMAP (n_neighbors = 10, init = pca, MN_ratio = 0.5, FP_ratio = 2), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5), (r) PaCMAP (n_neighbors = 5, init = pca, MN_ratio = 0.1, FP_ratio = 1), and (s) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.1, FP_ratio = 1). Is there a best representation of the original data or are they all providing  equivalent information?
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

  - Summarise contributions
  - Explain where it is expected or not expected to work, eg higher dimensional relationships
  - What might be useful enhancements

>>>>

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "none", 
     legend.title = element_text(size=5), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
interior_annotation <- function(label) {
  annotation_custom(grid::textGrob(label = label,
      x = unit(0.92, "npc"), y = unit(0.92, "npc"),
      gp = grid::gpar(cex = 1, col="grey70")))
}
```

@@

```{r}
# label: read-pbmc-nldr
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
clr_choice <- "#0077A3"
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")

nldr1 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("a")

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")

nldr2 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("b")

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.99.rds")
nldr3 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("c")

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")

nldr4 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("d")

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")

nldr5 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("e")

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
nldr6 <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("f")

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
nldr7 <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("g")

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
nldr8 <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("h")
```

```{r}
#| label: NLDR-variety
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
# 2D layouts from different NLDR techniques and different hyperparameter choices applied for the PBMC3k dataset: (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (d) UMAP (n_neighbors = 15, min_dist = 0.03), (e) UMAP (n_neighbors = 84, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (h) tSNE (perplexity = 92), (i) PHATE (knn = 2), (j) PHATE (knn = 5), (k) PHATE (knn = 10), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (m) TriMAP (n_inliers = 30, n_outliers = 4, n_random = 10), (n) TriMAP (n_inliers = 5, n_outliers = 2, n_random = 2), (p) PaCMAP (n_neighbors = 10, init = pca, MN_ratio = 0.5, FP_ratio = 2), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5), (r) PaCMAP (n_neighbors = 5, init = pca, MN_ratio = 0.1, FP_ratio = 1), and (s) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.1, FP_ratio = 1). Is there a best representation of the original data or are they all providing  equivalent information?
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

  - Summarise contributions
  - Explain where it is expected or not expected to work, eg higher dimensional relationships
  - Diagnostic app to explore differences in distances
  - What might be useful enhancements

### 2024-03-28 paper.tex
done with first image

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diffnldrpbmc-1.pdf}

@@

\caption{\label{fig-diffnldrpbmc}Six different NLDR representations of
the same data. Different techniques and different parameter choices are
used. Which is the most accurate representation of the structure in high
dimensions?}

@@

\begin{itemize}
\tightlist
\item
  Summarise contributions
\item
  Explain where it is expected or not expected to work, eg higher
  dimensional relationships
\item
  What might be useful enhancements
\end{itemize}

>>>>

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/NLDR-variety-1.pdf}

@@

\caption{Six different NLDR representations of the same data. Different
techniques and different parameter choices are used. Which is the most
accurate representation of the structure in high dimensions?}

@@

\begin{itemize}
\tightlist
\item
  Summarise contributions
\item
  Explain where it is expected or not expected to work, eg higher
  dimensional relationships
\item
  Diagnostic app to explore differences in distances
\item
  What might be useful enhancements
\end{itemize}
